---
title: "Class 10: Halloween mini project"
author: "Chaewoo Chun (A18548909)"
format: pdf
toc: true
---

## Background

As it is nearly halloween and the half way point in the quarter let's do a mini project to help us figure our the best candy!

Our come from the 538 website and is available as a CSV file:

## Data Import

```{r}
candy<- read.csv("candy-data.csv", row.names =1)
head(candy)
flextable::flextable(head(candy,10))
```

> Q1.
> How many different candy types are in this dataset?

```{r}
nrow(candy)
```

there are 85 different types of candies.

```{r}
candy|>
  nrow()
```

```{r}
 library(tidyverse)
candy %>%
  nrow()
```

> Q2.
> How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity ==1)
```

There are 38 of fruity candy in the dataset

> Q3.
> What is your favorite candy in the dataset and what is it’s winpercent value?

my favorite candy is Snickers!

```{r}
candy["Snickers", ]$winpercent
```

There are 76.67% of Snickers

```{r}
library(dplyr)

candy|> filter (rownames(candy)=="Snickers")|>
  select(winpercent)
```

> Q4.
> What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5.
> What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

## Quick overview of the dataset

> Q6.
> Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

```{r}
library(skimr)
skim(candy)
```

The winpercent is on a 0-100 scale the rest are 0-1 scale.

> Q7.
> What do you think a zero and one represent for the candy\$chocolate column?

```{r}
unique(candy$chocolate)
```

In candy\$chocolate, 1 means the candy contains chocolate, 0 means the candy does not contain chocolate.

> Q8.
> Plot a histogram of winpercent values

```{r}
hist(candy$winpercent,
     main = "Distribution of Win Percentages",
     xlab = "Win Percent",
     col = "skyblue", border = "white")
```

```{r}
library(ggplot2)
ggplot(candy, aes(x = winpercent)) +
  geom_histogram(bins= 20, fill = "skyblue", color = "white") +
  labs(title = "Distribution of Win Percentages", x = "Win Percent", y = "Count")
```

> Q9.
> Is the distribution of winpercent values symmetrical?

No — the distribution is is not symmetrical

> Q10.
> Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
median(candy$winpercent)
```

```{r}
summary(candy$winpercent)
```

The mean is about 50%, but the median is below 50%, so the center of the distribution is slightly below 50%.
That means most candies are less popular than average.

> Q11.
> On average is chocolate candy higher or lower ranked than fruit candy?

On average, chocolate candies have a higher winpercent (60.9%) than fruity candies (44.1%), meaning chocolate candies are more popular overall.

```{r}
#1. Find all chocolate candy in the dataset
#2. Find their winpercent values
#3. Calculate the mean of these values

#4-6. Do the same for fruity candy
#7. Compare mean winpercents of chocolate vs fruity
#8 Pick the highest as the winner

choc.inds <- candy$chocolate==1
choc.win <- candy[choc.inds, ]$winpercent
choc.mean <- mean(choc.win)
choc.mean
```

```{r}
mean(candy[candy$chocolate ==1,]$winpercent)
mean(candy[candy$fruity==1,]$winpercent)
```

```{r}
fruit.ind <- candy$fruity==1
fruit.win <- candy[fruit.ind,]$winpercent
fruit.mean <- mean(fruit.win)
fruit.mean
```

> Q12.
> Is this difference statistically significant?

```{r}
t.test(choc.win,fruit.win)
```

Chocolate candies are significantly more popular than fruity candies.
Chocolate candies have a mean winpercent of 60.9%, while fruity candies average 44.1%.
The p-value (2.87 × 10\^-8) indicates a highly significant difference.
The 95% confidence interval (11.45–22.16) shows that chocolate candies outperform fruity candies by roughly 11–22 percentage points on average.

> Q13.
> What are the five least liked candy types in this set?

```{r}
candy |>
  arrange(winpercent) |>
  head(5)
```

```{r}
x<- c(5,1,10,4)
#sort(x)
order(x)
#(candy$winpercent)
```

```{r}
order.ind <- order(candy$winpercent)
head(candy[order.ind,], 5)
```

> Q14.
> What are the top 5 all time favorite candy types out of this set?

```{r}
candy |>
  arrange(-winpercent) |>
  head(5)
```

> Q15.
> Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy)+
  aes(winpercent, rownames(candy))+
  geom_col()
```

> Q16.
> This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy)+
  aes(x=winpercent, y=reorder(rownames(candy), winpercent))+
  geom_col()
```

Add some color based on the "type of candy"

```{r}
my_cols <-rep("pink",nrow(candy))
my_cols[as.logical(candy$chocolate)] <- "lightblue"
my_cols[as.logical(candy$fruity)]<- "yellow"
my_cols[as.logical(candy$bar)] <- "orange"
my_cols
```

```{r}
ggplot(candy)+
  aes(x=winpercent, y=reorder(rownames(candy), winpercent))+
  geom_col(fill= my_cols)
```

> Q17.
> What is the worst ranked chocolate candy?

Nik L Rip

> Q18.
> What is the best ranked fruity candy?

Reeses Peanut Butter cup

## Winpercent and Pricepercent

A plot with both variables/columns winpercent and pricepercent

```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

> Q19.
> Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese’s Peanut Butter Cup offers the best value for money — it has one of the highest winpercents while not being among the most expensive.

> Q20.
> What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Nik L Nip, Nestle Smarties,Ring Pop, Hershey's Krackel, Hershey's Milk Chocolate are the top 5 most expensive and

Nik L Nip is the most least popular candy (22.44%)

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head(candy[ord, c(11,12)], n=5)
```

## Exploring the correlation structure

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

> Q22.
> Examining this plot what two variables are anti-correlated (i.e. have minus values)?

chocolate and fruity — when a candy is chocolaty, it’s usually not fruity, and vice versa.
chocolate and fruity are most anti-correlated.

> Q23.
> Similarly, what two variables are most positively correlated?

chocolate and bar are most positively correlated.

## Pricipal Component Analysis

The function to use is called `prcomp()` with an optional `scale=T/F` argument.

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

Our main PCA result figure

```{r}
ggplot(pca$x)+
  aes(PC1,PC2,
      label=rownames(pca$x))+
  geom_point(col =my_cols)+
  geom_text_repel(col=my_cols,max.overlaps =7)
```

We should also examine the variable "loadings" or contributions of the original variables to the new PCs.

```{r}
 ggplot(pca$rotation)+
  aes(PC1, rownames(pca$rotation))+
  geom_col()
```

```{r}
p<- ggplot(pca$x)+
  aes(PC1,PC2,
      label=rownames(pca$x))+
  geom_point(col =my_cols)+
  geom_text_repel(col=my_cols,max.overlaps =7)

```

Interactive plots that can be zoomed on and "brushed ocer an by made with the **plotly** package. It's output is interactive and will not render to PDF.

```{r}
library(plotly)
 
#plotly(p)
```

> Q24.
> What original variables are picked up strongly by PC1 in the positive direction?
> Do these make sense to you?

PC1 separates fruity/pluribus candies (positive) from chocolate/bar candies (negative).

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```
