---
title: "Class5: Data Viz with ggplot"
author: "Chaewoo (A18548909)"
format: pdf
---

Today we are exploring the **ggplot** package and how to make a nice figure in R.

There are lots of ways to make figures and plot in R. These include:

- so called "base" R
- and add on packages like **ggplot2**

Here is a simple "base" R plot.

```{r}
head(cars)
```

We can simply pass to the 'plot()' function.

```{r}
plot(cars)
```

 > Key-point : Base R is quick but not so nice looking in some folks eyes
 
 Let's see how we can plot this with **ggplot2**...

1st I need to install this add-on package. For this we use the 'install.packages()' function. - **WE DO THIS IN THR CONSOLE, NOT out report**. This is a one time only deal.

2nd We need to load the package with the 'library()' function every time we want to use it.

```{r}
library(ggplot2)
ggplot(cars)
```
 
 Every ggplot is composed of at least 3 layers:
 
 - **data**(i.e a data.frame with the things you want to plot),
 - aesthetics**aes()** that map the colums of data to your plot features (i.e aesthetics)
 -geoms like **geom_point()** that srt how the plot appears
 
```{r}
ggplot(cars)+
  aes(x=speed, y=dist)+
  geom_point()
```

>For simple "canned" gaphs base R is quicker but as things get more custom and elaborate then ggplot out...

Let's add more layers to out ggplot

Add a line showing the relationship betwen x and y
Add a title
Add a custom axis labels "Speed(MPH)" and "Distance(ft)"
Change the theme...

```{r}
ggplot(cars)+
  aes(x=speed, y=dist)+
  geom_point() + geom_smooth(method="lm", se=FALSE)+
 labs(title="Silly plot of Speed vs Stopping distance", x ="Speed(MPH)", y="Distance (ft)") + theme_bw()
```

#Going further

Read some gene expression data

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)

head(genes,5)
```

 >Q1. How many genes are in this wee dataset?
 
```{r}
nrow(genes)
```
 
 >Q2. How many "up" regulated genes are there?
 
```{r}
sum( genes$State =="up")
```
 
 A useful function for counting up occurances of things in a vector is the 'table()' function.
```{r}
table(genes$State)
```
 
 Make a v1 figure
 
```{r}
p<- ggplot(genes) +
  aes(x= Condition1,
     y= Condition2, col= State)+ 
  geom_point() 

p
```
 
```{r}
p+ scale_colour_manual(values = c("up" = "pink", "down" = "yellow", "unchanging" = "lightblue")) +labs(title="Gene expression changes upon drug treatment",x= "Control(no drug)" , y="Drug Treatment")+ theme_bw()
```
 
 
##More Plotting 

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

Lets have a wee peak
 
```{r}
head(gapminder, 3) 
```
 
```{r}
tail(gapminder, 3)
```
 
 >Q4. How many different county values are in this dataset?
 
```{r}
length(table(gapminder$country))
```
 >Q5. How many different continet values are in this dataset.
 
```{r}
length(table(gapminder$continent))
```

```{r}
unique(gapminder$continent)
```
 
```{r}
ggplot(gapminder) +
  aes(gdpPercap,lifeExp, col= continent) + 
  geom_point() 
```
 
```{r}
ggplot(gapminder) +
  aes(gdpPercap,lifeExp, col= continent, label=country) + 
  geom_point()+
  geom_text()
```

I can use the **ggrepel** package to make more sensivle labels here.


i want a seperate pannel per continent

```{r}
ggplot(gapminder) +
  aes(gdpPercap,lifeExp, col= continent, label=country) + 
  geom_point()+
  facet_wrap(~continent)
```

Layered Grammar of Graphics: ggplot2 uses a consistent, layered approach where you build plots by adding layers for data, aesthetics, and geometric objects. This makes complex plots easier to construct and modify 
1
, 
3
, 
2
.
Publication-Quality Defaults: ggplot2 produces attractive, publication-ready figures with sensible defaults, reducing the need for manual tweaking that is often required in base R 
1
, 
3
, 
2
.
Declarative Syntax: You specify what you want to see (data mappings, aesthetics, geoms) rather than how to draw each element, making code more readable and maintainable 
1
, 
3
, 
2
.
Customization and Extensibility: It is easier to customize and extend plots (e.g., adding color, themes, labels) using a consistent syntax, whereas base R often requires different functions and arguments for each plot type 
1
, 
3
, 
2
.
Reproducibility: ggplot2 code is scriptable and reproducible, allowing you to recreate and share plots easily 