---
title: "Week 8 Lab session"
author: "Chaewoo Chun (A18548909)"
format: pdf
---

# Section 1. proportion og G/G in a population

Downloaded a CSV file from Ensemble <https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39810234-39980235;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel>

 Here we read this CSV file
 
```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```
 
```{r}
table(mxl$Genotype..forward.strand.)
```
 
```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl) *100
```
 
 Now let's look at a different population. I picked the GBR.
 
```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(gbr)
```
 
 
 Find proportion of G|G
 
```{r}
table(gbr$Genotype..forward.strand.) /nrow(gbr)*100
```
 
 This variant that is associated with childhood asthma is more frequent in the GBR population than the MKL population.
 
 Lets dig into this further.
 
## Section 4: Population Scale Analysis
 
 One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
library(ggplot2)
```

Lets make a boxplot

```{r}
ggplot(expr) + aes(geno, exp, fill=geno) + 
  geom_boxplot(notch=TRUE)
```







 